On change of priors in one-vs-one classification methods

### 2025-05-08 coupling.tex
Fix proof of Proposition 3

For any sample, its result is identical to the arboreal method where the underlying tree forms a star. This method is clearly the optimal one with respect to Hinton's minimality criterion, because it ignores the output of any classifier not trained on the true class. It is not a true classification method, however, because it ``peeks'' at the label. Its main utility is to provide a way to benchmark other methods with respect to Hinton's minimality.

@@

A very helpful aid to study Bayes covariant methods is property $BC_{ij}$ (the notation stands for \emph{Bayes covariant when restricted to classes} $i$ and $j$). A coupling method $\boldsymbol{v}$ satisfies $BC_{ij}$ if for any change of priors $\pi$ and any  pairwise data $\boldsymbol{R}$ the odds of classes $C_i$ to $C_j$ predicted by $\boldsymbol{v}(\boldsymbol{R}^\pi)$ and $\bigl(\boldsymbol{v}(\boldsymbol{R})\bigr)^\pi$ are equal. Using  notation for odds from Section \ref{sec:one-vs-one}, one has
\begin{align}
	\alpha_{ij} \bigl(\boldsymbol{v}(\boldsymbol{R}^\pi) \bigr) = \alpha_{ij} \bigl(\boldsymbol{v}(\boldsymbol{R})^\pi \bigr)
\end{align}

@@

%	
%	
%Let $\pi$ be a change of priors from $\Pi = (P_1, \ldots, P_K)$ to $(P'_1, \ldots, P'_K)$. We need to verify $\boldsymbol{v}(\boldsymbol{R}^\pi) =\bigl( \boldsymbol{v}(\boldsymbol{R})\bigr)$ for any pairwise matrix $\boldsymbol{R}$. |We will rephrase this in an equivalent way. Let 
%\begin{align}
%	\begin{split}
%	\boldsymbol{v}(\boldsymbol{R}^\pi) &= (v'_1, \ldots, v'_K) \\
%	\boldsymbol{v}(\boldsymbol{R}) &= (v_1, \ldots, v_K)
%	\end{split}
%\end{align}
%Suppose $\pi$ is a change of priors from $\Pi = (P_1, \ldots, P_K)$ to $\Pi' = (P'_1, \ldots, P'_K)$. If the prediction given $\Pi$ is $(t_1, \ldots, t_K)$ then we need to prove that
%\begin{align*}
%\boldsymbol{t}(R^\pi) = \bigl(t_1 \frac{P'_1}{P_1}, \ldots, t_K \frac{P'_K}{P_K}\bigr)
%\end{align*}
%	
%	We can proceed by induction. The statement is clear when $K=2$. Suppose the proposition is true for an integers $<K$ and we are given a spanning tree $T$ on $K$ vertices. Any tree has a vertex of degree one. We may assume that in the case of $T$ it is the vertex corresponding to $C_1$ and the unique edge incident to $C_1$ is $C_1C_2$.
%	
%	 Denote by $\boldsymbol{t}$ the prediction corresponding to $T$ and by $\boldsymbol{t}'$ the prediction corresponding to the subgraph without the vertex $C_1$. Let us write 
%	\begin{align*}
%	\boldsymbol{t}(x) &= (t_1, t_2, \ldots, t_K) \\
%	\boldsymbol{t'}(x) &= ( t'_2, \ldots, t'_K) 
%\end{align*}
%These vectors are related as follows:
%\begin{align}
%\begin{split}
%\frac{t_1}{t_1+t_2} &= r_{12} \\
%(t_2, \ldots, t_K) &\propto (t'_2, \ldots, t'_{K})
%\end{split} \label{eq:ind1}
%\end{align}
%
%Now consider the effect of change of priors $\Pi = (P_1, \ldots, P_K)$ to $\Pi' = (P'_1, \ldots, P'_K)$. Applying \eqref{eq:changePrior} to the multi-class prediction yields a vector 
%$$
%\boldsymbol{t}^{\Pi\rightarrow \Pi'} \propto (t_1 \frac{P'_1}{P_1}, \ldots, t_K \frac{P'_K}{P_K}).
%$$
%Now consider what happens when we apply \eqref{eq:changePrior} before coupling. By induction assumption the method $\boldsymbol{t}'$ is Bayes covariant and thus applying coupling \emph{after} change of priors yields
%$$
%\boldsymbol{t}'^{\Pi\rightarrow \Pi'} \propto (t_2 \frac{P'_2}{P_2}, \ldots, t_K \frac{P'_K}{P_K})
%$$

@@

Given several Bayes covariant methods one can form  their ``linear combination'' - an ensemble which is also Bayes covariant. Let us make it explicit.

Suppose real numbers $a_1, \ldots,a_M$ are real numbers and $\boldsymbol{v}_1, \ldots, \boldsymbol{v}_M$ are probabibilistic multi-class classification methods satisfying Assumption \ref{ass:1}. We define their linear combination $\bigoplus_i a_m \boldsymbol{v}_m$ as the classification method that for any $x$ yields the probability distribution on $K$ classes that satisfies
\begin{align*}
 	\bigl(a_1 \boldsymbol{v}_1 \oplus \ldots \oplus a_M \boldsymbol{v}_M\bigr)(x) \propto \boldsymbol{v}_1^{a_1}(x) \odot \cdots \odot \boldsymbol{v}_M^{a_M}(x),
\end{align*}
where the symbol $\odot$ denotes componentwise multiplication, and exponentials are computed component-wise. With this definition we can state the following result.

\begin{prop}
Suppose the coupling methods $\boldsymbol{v}_1, \ldots, \boldsymbol{v}_M$ are Bayes covariant. If $\sum a_m = 1$, then $\bigoplus_m a_m \boldsymbol{v}_m$ is also a Bayes covariant coupling method.
\end{prop}

\begin{proof}
We will be using property $BC_{ij}$ from the proof of the previous proposition. It is sufficient to show that $BC_{ij}$ holds for any $i\not= j$.

Let $\pi$ be a change of priors from $(P_1, \ldots, P_K)$  to $(P'_1, \ldots, P'_K)$. Let $o_m$ denotes the odds of classes $C_i$ and $C_j$ predicted by coupling method $\boldsymbol{v}_m$. Then the odds of $C_i$ and $C_j$ predicted by $(\bigoplus_i a_i \boldsymbol{v}_i)^\pi$ are 
\begin{align}
	\biggl(\bigoplus_m a_m \boldsymbol{v}_m\biggr)^\pi(\boldsymbol{R}) & = \frac{P'_i}{P_i} \frac{P_j}{P'_j} \biggl(\bigoplus_m a_m \boldsymbol{v}_m\biggr)(\boldsymbol{R}) \\
\end{align}
so the odds are

\begin{align}
	&= \frac{P'_i}{P_i} \frac{P_j}{P'_j} \prod_m o_m^{a_m}
\end{align}

Let us now use Bayes covariance of $\boldsymbol{v}_i$. 
We have
\begin{align}
	\biggl(\bigoplus_m a_m \boldsymbol{v}_m\biggr)(\boldsymbol{R^\pi}) & =  \biggl(\bigoplus_m a_m \boldsymbol{v}_m^\pi\biggr)(\boldsymbol{R}) \\
%	&= \frac{P'_i}{P_i} \frac{P_j}{P'_j} \prod_m o_m^{a_m}
\end{align}

It follows that the odds of $C_i$ to $C_j$ for $\bigoplus_m a_m \boldsymbol{v}_m$ are 
\begin{align}
\biggl(\frac{P'_i}{P_i} \frac{P_j}{P'_j} \biggr)^{\sum a_m} \prod_m o_m^{a_m}.
\end{align}
Clearly, under the assumption $\sum_m a_m = 1$, the classifier $\bigoplus_m a_m \boldsymbol{v}_m$ satisfies $BC_{ij}$ and it follows that it is Bayes covariant.
\end{proof}

>>>>

For any one sample belonging to class $C_t$, its result is identical to the arboreal method where the underlying tree forms a star with center at $C_t$. This method is clearly the optimal one with respect to Hinton's minimality criterion, because it ignores the output of any classifier not trained on the true class. It is not a true classification method, however, because it ``peeks'' at the label. Its main utility is to provide a way to benchmark other methods with respect to Hinton's minimality.

@@

A very helpful aid to study Bayes covariant methods is property $BC_{ij}$ (the notation stands for \emph{Bayes covariant when restricted to classes} $i$ and $j$). A coupling method $\boldsymbol{v}$ satisfies $BC_{ij}$ if for any change of priors $\pi$ and any  pairwise data $\boldsymbol{R}$ the odds of classes $C_i$ to $C_j$ predicted by $\boldsymbol{v}(\boldsymbol{R}^\pi)$ and $\bigl(\boldsymbol{v}(\boldsymbol{R})\bigr)^\pi$ are equal. Using earlier notation for odds (see \eqref{eq:alpha}), one has
\begin{align}
	\alpha_{ij} \bigl(\boldsymbol{v}(\boldsymbol{R}^\pi) \bigr) = \alpha_{ij} \bigl(\boldsymbol{v}(\boldsymbol{R})^\pi \bigr)
\end{align}

@@

@@

Given several Bayes covariant methods one can form  their ``linear combination'' - an ensemble which is also Bayes covariant. Let us make the construction explicit.

Suppose first that real numbers $a_1, \ldots,a_M$ are real numbers and $\boldsymbol{v}_1, \ldots, \boldsymbol{v}_M$ are probabibilistic multi-class classification methods satisfying Assumption \ref{ass:2}. We define their linear combination $\bigoplus_i a_m \boldsymbol{v}_m$ as the classification method that for any $x$ yields the probability distribution on $K$ classes that satisfies
\begin{align*}
 	\bigl(a_1 \boldsymbol{v}_1 \oplus \ldots \oplus a_M \boldsymbol{v}_M\bigr)(x) \propto \boldsymbol{v}_1^{a_1}(x) \odot \cdots \odot \boldsymbol{v}_M^{a_M}(x),
\end{align*}
where the symbol $\odot$ denotes componentwise multiplication, and exponentials are computed component-wise. With this definition we can state the following result.

\begin{prop}
Suppose the coupling methods $\boldsymbol{v}_1, \ldots, \boldsymbol{v}_M$ are Bayes covariant. If $\sum a_m = 1$, then $\boldsymbol{e} = \bigoplus_m a_m \boldsymbol{v}_m$ is also a Bayes covariant coupling method.
\end{prop}

\begin{proof}
It is sufficient to show that $BC_{ij}$ holds for any $i\not= j$.

Let $\pi$ be a change of priors from $(P_1, \ldots, P_K)$  to $(P'_1, \ldots, P'_K)$. Let $o_m$ denotes the odds of classes $C_i$ and $C_j$ predicted by coupling method $\boldsymbol{v}_m$ i.e.
$$
o_m = \alpha_{ij}(\boldsymbol{v}_m(\boldsymbol{R})).
$$

It follows from the definition of linear combination of coupling methods that the odds of $C_i$ to $C_j$ for $\boldsymbol{e}$ are
\begin{align}
\prod_m o_m^{a_m}.
\end{align}
It follows that the odds of $\boldsymbol{e}^\pi$ are 
\begin{align}
\frac{P'_i}{P_i} \frac{P_j}{P'_j} \prod_m o_m^{a_m}.
\end{align}

Let us now use Bayes covariance of $\boldsymbol{v}_i$. 
We have
\begin{align}
	\biggl(\bigoplus_m a_m \boldsymbol{v}_m\biggr)(\boldsymbol{R^\pi}) & \propto 
	\boldsymbol{v}_1^{a_1}(\boldsymbol{R}^\pi) \odot \cdots \odot \boldsymbol{v}_M^{a_M}(\boldsymbol{R}^\pi) =  \bigodot_m  (\boldsymbol{v}_m^\pi(\boldsymbol{R}))^{a_m} \\
%	&= \frac{P'_i}{P_i} \frac{P_j}{P'_j} \prod_m o_m^{a_m}
\end{align}

It follows that the odds of $C_i$ to $C_j$ for $\bigoplus_m a_m \boldsymbol{v}_m$ are 
\begin{align}
\biggl(\frac{P'_i}{P_i} \frac{P_j}{P'_j} \biggr)^{\sum a_m} \prod_m o_m^{a_m}.
\end{align}
Clearly, under the assumption $\sum_m a_m = 1$, the classifier $\bigoplus_m a_m \boldsymbol{v}_m$ satisfies $BC_{ij}$. Since this holds for any $i\not=j$,  it follows that $\boldsymbol{e}$ is Bayes covariant.
\end{proof}
